---
## Compile with
## Rscript -e "rmarkdown::render('presentation.Rmd')"
title: Missing Data
subtitle: https://github.com/martingerdin/missing-data-presentation
author: Martin Gerdin Wärnberg
date: 2022-03-21
output: beamer_presentation
bibliography: bibliography.bib
csl: chicago-author-date.csl
---

# Reviewer comment

> As the authors state, half the patients were excluded as a result of
> missing data. Did these patients differ from the patients who were
> included with regard to age, gender mechanism of injury, severity of
> TBI, etc?

# What does the reviewer want?

```{r, cache=TRUE, include=FALSE}
data <- rio::import("https://raw.githubusercontent.com/titco/titco-I/master/titco-I-full-dataset-v1.csv")
vars <- c("age", "sex", "sbp_1", "rr_1", "gcs_t_1", "died")
strata <- "complete"
table.data <- data[, vars]
table.data$age <- as.numeric(table.data$age)
names <- c("Age", "Sex", "SBP", "RR", "GCS", "Died")
names(table.data) <- names
nonnormal <- names[sapply(table.data, is.numeric)]
table.data$complete <- complete.cases(table.data)
table.data$complete <- factor(table.data$complete,
                              levels = c(TRUE, FALSE),
                              labels = c("Complete", "Incomplete"))
table <- print(tableone::CreateTableOne(vars = names,
                                        data = table.data,
                                        strata = strata,
                                        test = FALSE),
               nonnormal = nonnormal,
               showAllLevels = TRUE,
               catDigits = 0,
               contDigits = 0)
attr(table, which = "dimnames")$`Stratified by complete`[1] <- "Level"
```

```{r, echo=FALSE, results="asis"}
knitr::kable(table, caption = "Comparison of complete and incomplete observations")
```
Abbreviations: GCS Glasgow Coma Scale, RR Respiratory Rate, SBP Systolic Blood Pressure

# What is missing data?

- Data that was never collected
- Data that was not available
- Common in all study types

# Why is missing data problematic?

- Lower statistical power
- Loss of key subgroups
- Biased or inaccurate estimates
- Increased analysis complexity

# What can be done about missing data?

- Avoid it
- Manage it appropriately
- Conduct sensitivity analyses

# What are the different missing data mechanisms?

- Missing completely at random (MCAR)
- Missing at random (MAR)
- Missing not at random (MNAR)

# What is MCAR?

> **Observations of all subjects are equally likely to be
> missing**. That is, there are no systematic differences between
> subjects with observed and unobserved values meaning that the
> observed values can be treated as a random sample of the
> population. For example, echocardiographic measurements might be
> missing due to sporadic ultrasound malfunction.
>
> -- [@Papageorgiou2018]

# What is MAR?

> **The likelihood of a value to be missing depends on other, observed
> variables**. Hence, any systematic difference between missing and
> observed values can be attributed to observed data. That is, the
> relationships observed in the data at hand can be utilized to
> ‘recover’ the missing data. For example, missing echocardiographic
> measurements might be more normal than the observed ones because
> younger patients are more likely to miss an appointment.
> 
> -- [@Papageorgiou2018] 

# What is MNAR?

> **The likelihood to be missing depends on the (unobserved) value
> itself**, and thus, systematic differences between the missing and
> the observed values remain, even after accounting for all other
> available information. In other words, there is extra information
> associated with the missing data that cannot be recovered by
> utilizing the relationships observed in the data. For example,
> missing echocardiographic measurements might be worse than the
> observed ones because patients with severe valve disease are more
> likely to miss a clinic visit because they are unable to visit the
> hospital.
> 
> -- [@Papageorgiou2018] 

# How can we tell different mechanisms apart?

- We can't use only the data
- Requires knowledge and reasoning about how the data was generated

# How can missing data be managed?

- Ignored
- Complete case analysis
- Mean imputation
- Single or deterministic imputation
- Stochastic imputation
- Multiple imputation
- Sensitivity analyses (subgroups, best-worst case scenarios)
- Other ways

# How are missing data mechanisms and methods related?

|                        | MCAR          | MAR       | MNAR            |
|------------------------|---------------|-----------|-----------------|
| Ignore                 | Loss of power | Bad idea  | Bad idea        |
| Complete case analysis | Loss of power | Biased    | Biased          |
| Mean imputation        | Biased        | Biased    | Biased          |
| Single imputation      | Biased        | Biased    | Biased          |
| Stochastic imputation  | Probably okay | Biased    | Biased          |
| Multiple imputation    | Okay          | Okay      | Probably biased |
| Sensitivity analyses   | Not needed    | Good idea | Good idea       |

# Why is ignoring missing data a bad idea?

Treating incomplete data as complete

- Loss of control
- Loss of denominator
- Incomparable groups
- Unknown precision

# How can we interpret results based on ignored missing data?

```{r, cache=TRUE, include=FALSE}
get_results <- function(x, y, data) {
    model.data <- data.frame(y = data[, y], data[, x])
    model <- glm(y ~ ., data = model.data, family = binomial())
    pe <- c(model$coefficients[-1])
    ci <- as.matrix(confint(model)[-1, ])
    if (length(x) == 1)
        ci <- t(ci)
    v <- round(exp(cbind(pe, ci)), 3)
    pv <- c(n = nrow(model$model), est = paste0(v[, 1], " (", v[, 2], "-", v[, 3], ")"))
    return(pv)
}
models.data <- table.data
models.data$Died <- as.numeric(models.data$Died == "Yes")
create_table <- function(data, include.n = TRUE) {
    variables <- setNames(nm = c("Age", "Sex", "SBP", "RR", "GCS"))
    unadjusted.results <- do.call(rbind, lapply(variables, get_results, y = "Died", data = data))
    adjusted.results <- get_results(variables, "Died", data = data)
    adjusted.n <- adjusted.results["n"]
    if (!include.n)
        unadjusted.results <- unadjusted.results[, -1]
    combined <- as.data.frame(cbind(unadjusted.results, adjusted.results[-1]))
    if (include.n)
        names(combined) <- c("n^a^", "Unadjusted OR (95% CI)", "Adjusted OR (95% CI)^b^")
    if (!include.n)
        names(combined) <- c("Unadjusted OR (95% CI)", "Adjusted OR (95% CI)")
    attr(combined, "adjusted.n") <- adjusted.n
    return(combined)
}
ignored.table <- create_table(models.data)
```

```{r, echo=FALSE, results="asis"}
knitr::kable(ignored.table, caption = "Unadjusted and adjusted associations between variables and mortality")
```
Abbreviations: GCS Glasgow Coma Scale, OR Odds Ratio, RR Respiratory Rate, SBP Systolic Blood Pressure

Notes: ^a^The number of complete observations per variable. ^b^The number of complete cases used in the adjusted model was `r attr(ignored.table, "adjusted.n")`.

# How can we interpret results based on complete data (complete case analysis)?

```{r, include=FALSE}
complete.data <- models.data[complete.cases(models.data), ]
complete.table <- create_table(complete.data, include.n = FALSE)
```

```{r, echo=FALSE, results="asis"}
knitr::kable(complete.table, caption = paste0("Unadjusted and adjusted associations between variables and mortality (n = ", attr(complete.table, "adjusted.n"), ")"))
```

# How is missing data represented in different software?

| Software               | Missing data representation |
|------------------------|-----------------------------|
| Google Sheets (Excel?) |                             |
| STATA                  | ., .a-.z                    |
| SPSS                   | .                           |
| R                      | NA                          |

# How do software deal with missing data?

- Google Sheets (Excel?), STATA and SPSS will in most cases ignore
  missing data and use only the observed data to calculate some
  metric.
- R will in most cases return NA if you try to calculate some metric
  using data that includes missing values.
- For example, given this vector of systolic blood pressures: `120,
  90, 90, NA, 110`, to calculate a mean Google Sheets, STATA, and SPSS
  would return `102.5` whereas R would return `NA`.
- When doing some regression, STATA, SPSS, and R all default to a
  complete case analysis.

# What is mean imputation?

- Missing values are replaced with the mean of the observed data, for
example `120, 90, 90, NA, 110` would become `120, 90, 90, 102.5, 110`.n
-  Artificial reduction in variability.

# What is single or deterministic imputation?

- Missing values are replaced with the predicted scores from a regression equation.

\pause
```{r, echo=FALSE, results="asis"}
mini.data <- head(models.data)[-7]
mini.data$GCS[is.na(mini.data$GCS)] <- 11
knitr::kable(mini.data)
```

\pause
- Impute RR as:

$\bar{RR} = \beta_0 + \beta_1 Age + \beta_2 Sex + \beta_3 SBP + \beta_ 4 GCS + \beta_5 Died$

# What is single or deterministic imputation?

- Running that linear regression results in:

```{r, include=FALSE}
rr.model <- glm(RR ~ Age + Sex + SBP + GCS + Died, data = complete.data)
coefficients <- as.list(round(rr.model$coefficients, 2))
names(coefficients)[1] <- "b0"
rr1 <- round(predict(rr.model, newdata = mini.data[2, ]))
rr2 <- round(predict(rr.model, newdata = mini.data[5, ]))
attach(coefficients)
```

$\bar{RR} = `r b0` + `r Age`\cdot Age + `r SexMale`\cdot Sex + `r SBP`\cdot SBP + `r GCS`\cdot GCS + `r Died`\cdot Died$

\pause
- Meaning that for this patient:

```{r, echo=FALSE, results="asis"}
knitr::kable(mini.data[2, ], row.names = FALSE)
```

\pause
- We could estimate the missing RR as: 

$\bar{RR} = `r b0` + `r Age`\cdot 21 + `r SexMale`\cdot 1 + `r SBP`\cdot 110 + `r GCS`\cdot 11 + `r Died`\cdot 0 \approx 19$

# What is stochastic imputation?

- Improves on the single imputation by adding random noise.
  - Single imputation too perfect, amplifies associations (even under MCAR). 

\pause
- The modified regression equation:
$\bar{RR} = \beta_0 + \beta_1 Age + \beta_2 Sex + \beta_3 SBP + \beta_ 4 GCS + \beta_5 Died + \epsilon$

- Where $\epsilon$ is the noise term.

# What is multiple imputation?

> **The purpose of multiple imputation is to generate possible values
> for missing values, thus creating several "complete" sets of
> data**. Analytic procedures that work with multiple imputation
> datasets produce output for each "complete" dataset, plus pooled
> output that estimates what the results would have been if the
> original dataset had no missing values. These pooled results are
> generally more accurate than those provided by single imputation
> methods.
> 
> -- [@SPSS]

# What is multiple imputation?

- While less perfect than single imputation, stochastic imputation still does not incorporate the uncertainty associated with missing data.
- Multiple imputation uses stochastic imputation iteratively.
- Creates multiple complete datasets.
- Each imputed dataset is analysed using standard methods.
- The results from all analyses are pooled to create a combined estimate. 

# What is multiple imputation?

- You need to decide:
  - How many imputed datasets to create (M).
  - How each variable with missing data should be imputed.
  - What variables to include in the imputation model.
- The defaults:
  - 5 imputed datasets ($M = 5$). Some implementations use 20 ($M = 20$)
  - Logistic regression for binary variables, linear regression for
    continuous variables.
- More missing data, greater uncertainty, increase M (maybe same as % missingness).

# How is multiple imputation implemented?

- STATA: `mi`
- SPSS: `Analyze > Multiple Imputation > Impute Missing Data Values...`
- R: `mice`

# Some practical advice

- Unless technically inclined, use complete case analysis.
- Document the amount of missing values in each variable of interest (table).
- Document the number of observations (patients) before incomplete.
  cases were removed, and the number of observations remaining.
- Show how you went from the original data to the study sample (flowchart).
- Present missing data in the first paragraph in Results.
- Include a table (supplementary material?) comparing complete and incomplete observations.
- Acknowledge possible biases and reflect on these in Limitations.

# Missing values table

```{r, echo=FALSE, results="asis"}
missing.data <- as.data.frame(do.call(rbind, lapply(models.data[, -7], function(x) c(sum(is.na(x)), round(mean(is.na(x)) * 100, 1)))))
names(missing.data) <- c("n", "%")
knitr::kable(missing.data, caption = "Missing values per variable of interest")
```
Abbreviations: GCS Glasgow Coma Scale, RR Respiratory Rate, SBP Systolic Blood Pressure

# Flowchart

```{r, engine="dot", echo=FALSE, fig.cap="Study flowchart"}
digraph {
    node [ shape = rect, fontname = "Computer Modern" ]

    rankdir = "LR";
    splines = false;
    nodesep = 0.5;
    
    a [ label = "16 000 patients in original cohort"]
    a1 [ width = 0, shape = point, style = invis ]
    b [ label = "8 000 patients aged 15 or older years"]
    b1 [ width = 0, shape = point, style = invis ]
    c [ label = "5 000 patient had complete data\n and constituted the final sample"]
    e1 [ label = "8 000 patients excluded because\n they were younger than 15 years" ]
    e2 [ label = "3 000 patients excluded because\n they had missing data*:\n-21 patients had missing age\n-1800 patients had missing SBP\n-1200 patients had missing RR\n-400 patients had missing GCS"]

    a -> a1 [ dir = none ];
    a1 -> e1;
    a1 -> b;
    b -> b1 [ dir = none ];
    b1 -> e2;
    b1 -> c;

    {
	rank = same;
	a; a1; b; b1; c;
    }
}
```

Note: *The sum of missing values may exceed the total number of patients with incomplete data because some patients have missing values in multiple variables.

# Table comparing complete and incomplete observations

```{r, echo=FALSE, results="asis"}
knitr::kable(table, caption = "Comparison of complete and incomplete observations")
```
Abbreviations: GCS Glasgow Coma Scale, RR Respiratory Rate, SBP Systolic Blood Pressure


# References
